---
title: "Exercises Day 1 Part 2- Greesham Simon"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ## Load the data

Load the `gapminder` dataset.


```{r LoadData}
library(here)
gapminder <- read.csv(here("data/gapminder5.csv"), 
                      stringsAsFactors=FALSE)
```

## ## If Statement

Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

Hint: use the any function.


```{r IfStatement}
test_year<- 2002
if(any(gapminder$year == test_year)){
    cat("There are records from",test_year,"in the gapminder dataset.")
} else{
    cat("There aren't any records from",test_year,"in the gapminder dataset.")
}

test_year<- 2012
if(any(gapminder$year == test_year)){
    cat("There are records from",test_year,"in the gapminder dataset.")
} else{
    cat("There aren't any records from",test_year,"in the gapminder dataset.")
}
```
## ## Loop and If Statement

Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

Write a script that loops through the gapminder data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.


```{r LoopAndIfStatement}
mean_pop<- mean(gapminder$pop)

for(i in unique(gapminder$country)) {
    country_mean <- mean(gapminder$pop[gapminder$country==i])
    
    if (country_mean < mean_pop) {
        cat("\nMean life expectancy for",i," is",mean(gapminder$lifeExp[gapminder$country == i]),".")
    }
}

lowLim <- 50
highLim <-70

continents <- unique(gapminder$continent)

for(i in continents){
    tmp <- mean(gapminder$lifeExp[gapminder$continent==i])
    if (tmp< lowLim) {
        cat(sep="","\nContinent ", i," has a mean life expectancy below ", lowLim,"." )
    } else if (tmp > highLim){
         cat(sep="","\nContinent ", i," has a mean life expectancy above ", highLim,"." )
    } else {
         cat(sep="","\nContinent ", i," has a mean life expectancy between ", lowLim," and ", highLim, "." )
    }
}

```

## ## Exercise: Write Functions

Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset. Hint: Use mode() or class() to get the class of the data in each column. Remember that names() or colnames() returns the name of the columns in a dataset.

Create a function that given a vector will print the mean and the standard deviation of a vector, it will optionally also print the median. Hint: include an argument that takes a boolean (TRUE/FALSE) operator and then include an if statement.


```{r WriteFunctions}

#first function prints name of each column and class

colNamesClass<- function(df) {
    cols <-names(df)
    for (i in cols){
        cat("\n",i,":",class(df[[i]]))
    }
}

colNamesClass(gapminder)
    
#second function prints mean and standard deviation. It may print median if printMedian is set to True

vectorBasicInfo <- function(x, printMedian=FALSE){
    cat("Mean:",mean(x),"\nStandard Deviation:",sd(x))
    if (printMedian){
        cat("\nMedian:", median(x))
    }
}

vectorBasicInfo(gapminder$lifeExp)
vectorBasicInfo(gapminder$lifeExp, TRUE)

```

## ## Exercise: Analyzing the Relationship

Use what you???ve learned so far to answer the following questions using the gapminder dataset. Be sure to include some visualizations!

1. What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)

2. Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.


```{r Analyzing the Relationship}
#Problem 1
#look at initial plot
plot(x=gapminder$gdpPercap,y=gapminder$lifeExp)

#look at transformed plot (nat log)
plot(x=log(gapminder$gdpPercap),y=log(gapminder$lifeExp))

linModel<- lm(log(gapminder$lifeExp)~log(gapminder$gdpPercap))
summary(linModel)
plot(x=log(gapminder$gdpPercap),y=log(gapminder$lifeExp))
abline(a=linModel$coefficients[1], b=linModel$coefficients[2])

# the power model looks like a good fit between life expectancy and gdp per capita


#Problem 2
# identify the countries in the continent Americas
unique(gapminder$country[gapminder$continent=="Americas"])

#Create vector of North American countries
nAmerica <- c("Canada", "Costa Rica","Cuba","Dominican Republic","El Salvador","Guatemala","Hait","Honduras","Jamaica", "Mexica", "Nicaragua", "Panama", "Trinidad and Tobago", "United States")


#Create New Continent column in gapminder
gapminder$NewCont <- gapminder$continent

#Parses through each column and assigns N America to North American countries and South America to the other "Americas" countries
for (i in 1:length(gapminder$NewCont)){
    if(gapminder$NewCont[i] == "Americas"){
        if(gapminder$country[i] %in% nAmerica){
            gapminder$NewCont[i] <- "North America"
        } else{
            gapminder$NewCont[i] <- "South America"
        }
    }
    
}

continents_new <- unique(gapminder$NewCon)
gapminder$logLifeExp <- log(gapminder$lifeExp)
gapminder$loggdpPerCap <- log(gapminder$gdpPercap)

for (i in continents_new){
    lm_ex <- lm(gapminder$logLifeExp[gapminder$NewCont==i]~ gapminder$loggdpPerCap[gapminder$NewCont==i])
    
    cat("\n", i, ": intercept=", lm_ex$coefficients[1],", slope=", lm_ex$coefficients[2])
    
}


```
