---
title: "R Day 2 exercises"
author: "Greeham Simon"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: TRUE
---

```{r global_options, echo = FALSE, include = FALSE}
answers<-params$answers

knitr::opts_chunk$set(echo=answers, eval=answers,
                      warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

<!-- BEFORE YOU START, REMEMBER TO RENAME THE .RMD FILE AND SAVE IT AS day2_exercises_LastNameFirstName.Rmd AND THEN REMOVE THIS INSTRUCTION -->

## Load the data

Read both California energy datasets. Make sure the `datetime` variable is in an appropriate data type (i.e. not character). 

```{r}
library(dplyr)
library(data.table)
library(lubridate)
library(readr)
generation_dp <- read_csv(here::here("data/ca_energy_generation.csv"))
imports_dp <- read_csv(here::here("data/ca_energy_imports.csv"))
generation_dt <- as.data.table(generation_dp)
imports_dt <- as.data.table(imports_dp)
str(imports_dt)

```


## Merge and reshape the data

Merge the two datasets and then melt the resulting dataframe/datatable to make it tidy.

```{r}
#inner join between data sets

all_data <- melt(generation_dt[imports_dt, on = "datetime"], 
                          id.vars = "datetime",
                          variable.name = "source",
                          value.name = "output")



```


## Creating new variables

Create a series of new variables: 

1. `day`, which is the year-month-day, without the hour. The `lubridate` function `as_date` will do this.
2. `log_output`, which is the natural log of the output.
3. **Challenge**: `per_output`, which is the percent of daily output represented by each observation. You will need to use `group_by` and to create a new variable with the total output for the day. (Make sure to use `ungroup()` after this!)

Bonus: If you are using `dplyr`, try to do this all in one pipe!

```{r}

all_data[,day := as_date(datetime)]
all_data[,log_output:= log(output)]
all_data[,per_output:= output/sum(output, na.rm=T),by = day]

```


## Summarizing and analyzing data

1. Which source has the greatest mean output by hour? (Hint: Use the `dplyr` verb `arrange(desc(variable))` to order the data frame so that the largest value of `variable` is first. Don't use `desc` and it arranges in ascending order. The `data.table` function is `setorder`.) Which has the least?
2. Which source has the greatest mean output by day? Which has the least? (Do not include zero values.)
3. Which sources has the greatest variance in usage over the course of a dataset? Which has the least? (Do not include zero values.)

```{r}
#1
#stores mean of output by source in mean_hrly
mean_hrly <- all_data[,.(mean_hourly = mean(output)), by = source]
#displays max and mein mean hourly rate by source
mean_hrly[mean_hourly == max(mean_hourly) | mean_hourly == min(mean_hourly)]

#2
#stores mean by day and source
mean_daily <- all_data[output>0,.(mean_d = mean(output)), by = .(source,day)]
#prints out the max and min
mean_daily[mean_d == max(mean_d) | mean_d == min(mean_d)]

#3
#stores standard deviation of output by source
sd_hrly <- all_data[output>0,.(sd_hourly = sd(output)), by = source]
#displays max and mein sd  by source
sd_hrly[sd_hourly == max(sd_hourly) | sd_hourly == min(sd_hourly)]
```


## Analyzing renewable versus non-renewable energy sources

The dataset `regroup.csv` has information about which sources are considered renewable by the state of California. Use this dataset, along with yourdata manipulation skills, to explore the use of renewable and non-renewable sources. Annotate what your descisions for the analysis.

Hint: Use your merge skills to merge the CA energy data with the `regroup` data. Which variable should you join by?

```{r}
regroup_dt <- fread(here::here("data/ca_energy_regroup.csv"))
str(regroup_dt)
#merged the two data sets to have a new "group" variable
updated_all_data<- merge(all_data, regroup_dt, by.x= "source", by.y = "type")
updated_all_data[,.(mean_vals = mean(output)), by= group]
updated_all_data[,.(sd_vals = sd(output)), by= group]
```

